Lyman-$\alpha$ photons through rotating outflows

### 2018-09-05 paper/paper.tex
almost final version

\begin{abstract}
Outflows and rotation are two ubiquitous kinematic features in the gas
kinematics of galaxies. 
Here we introduce a semi-analytic model to quantify how rotating
outflows impact the morphology of the Lyman-$\alpha$ emission line.   
The model is contrasted against Monte Carlo radiative transfer
simulations of outflowing gas with additional solid body rotation.
We explore a range of neutral Hydrogen optical depth of
$10^5\leq\tauh\leq10^7$, rotational velocity $0\leq \vrot/\kms \leq
100$ and outflow velocity $0\leq \vout/\kms\leq 50$.  
We find three consequences of rotation.
First, it introduces a dependency with viewing angle; second it
produces a line broadening and third it increases the flux at the
line's center.
For all viewing angles, the semi-analytic model reproduces the
radiative transfer results for the line width and flux change at the
line's center within a $7\%$ and $50\%$ precision for an optical depth of
$\tauh=10^5$, respectively, and within $2\%$ and $1\%$ for an optical
depth of $\tauh=10^7$.
We also show that another, less precise, approximation is considering
that rotation can be represented by a gaussian smoothing with standard
deviation $\frac{1}{2} \vrot \sin \theta$, where $\theta$ is the angle between
the line-of-sight and the rotation axis. 
The semi-analytic model presented here is a convenient tool to
introduce rotational kinematics as a post-processing step of idealized
Monte Carlo simulations, providing a framework to interpret \lya
spectra in systems where rotation is expected, or directly measured
through kinematic maps.  
\end{abstract}

@@

However, the gas dynamics in Lyman Alpha Emitter (LAE) galaxies are
more complex than pure rotation.
In many observations the \lya line profile has a single peak
redwards from the line's center, in other cases there is a double peak
but the peak on the red side is stronger
\citep[e.g.][]{2010ApJ...717..289S,Erb14,Trainor16}.   
These features have been explained as the consequence of multiple
\lya photon scatterings through a homogeneous outflowing shell of
neutral Hydrogen
\citep{2006A&A...460..397V,Orsi12,2015ApJ...812..123G}.

@@

\begin{figure}
\centering
\includegraphics[width=0.45\textwidth]{muse_vs_simulations.pdf}

\caption{\textbf{Comparison between simulations and observations.}  
The observational results correspond to the MUSE-Wide Survey
\citep{2017A&A...606A..12H}. The skewness and standard deviation from
the simulations summarize the results presented in Figures
\ref{fig:standard_deviation} and \ref{fig:skewness}.
  \label{fig:muse}}
\end{figure}

@@

\subsection{\lya Kinematic Maps}
\label{subsec:kinematic}
Current observational facilities have the capability of spatially
resolving the extent of a LAE.
For instance \cite{Prescott14} presented observational results of a
Doppler shift when taking spectra at two opposite sides of a large
($\approx 80$kpc) LAE.
In more recent work \cite{2018MNRAS.473.3907A} mapped \lya emission
around a quasar.

@@

Although it is a good approximation to think the rotating spectra by a
sum of Doppler shifts, the peak of the spectra is also weighted by the
amount of mass with a given line-of-sight velocity. 
In this toy model the distance between the peaks of the
receding/approaching spectra is close to $\sim 25\ \kms$, which is
a fourth of the naively expected value of $2\vrot=100\ \kms$,
due to the fact that only a small fraction of the photons are emitted
at the extreme of the galaxy having the maximum rotational velocity of
$50\ \kms$.

Spectrographs like the Multi Unit Spectroscopic Explorer (MUSE) could
obtain kinematic information from large samples of LAEs to build
velocity maps in Ly$\alpha$.  
This could be a natural extension of the work reported by
\cite{Herenz2016} on the velocity maps of several LARS (Lyman Alpha
Reference Sample) galaxies.  
The interpretation of such data should take into account the insights
and trends we have presented in this paper.

\subsection{Comparison against the MUSE-Wide Survey}

In order to give the reader a feeling of the ranges of observational
data that could be explained by the models presented here, 
we perform a comparison against observational data from the MUSE-Wide
Survey (MWS) \citep{2017A&A...606A..12H}.

MWS provides public access to 831 spectra out of which 237 contain a
detected \lya line as dominant. 
The public dataset reports as the redshift of the galaxy the position
of the dominant peak.
We use this redshift to convert the spectra to a rest-frame and compute
the standard deviation and skewness statistics. 
Given the uncertainty on the galaxy redshift we do not compute the
valley/peak ratio. 
We do not try to perform a full fit to all the spectra
\citep[e.g.][]{2017A&A...608A.139G} and wish only to
know whether the region of parameter space explored in our RT
simulations shows any crossing with the results from the MWS or
shows any sign of atypical features.

Figure \ref{fig:muse} shows the results of this experiment. 
The small white circles correspond to the 217 \lya emitters in the MWS
observations; the large black circles to the results of the
simulations presented in this paper. 
This shows that the systems modeled in this paper actually present an
overlap with statistics derived from observations. 
The region with standard deviation $>200$\kms could also be explained
by systems with larger optical depths than we have used here. 
However, the systems with a positive skewness, that is a line
unbalanced towards the blue could be the sign of inflowing
kinematics.

%\subsection{Low optical thickness regimes}
%When located in the regimes of low optical thickness, \tauh $\lesssim
%4$,  the agreement of the radiative transfer simulation with the
%semi-analytical solution is not as accurate. In this regime, only
%numerical simulations can help to predict the shape of the \lya
%profiles. We ran a simulation with low \tauh and compared it with the
%semi-analytical model. As seen in Figure \ref{fig:low_tau}, the model
%does not fit the simulation. This creates physical restrictions on the
%input of our model.

%\begin{figure}
%	\centering
%	\includegraphics[width=0.48\textwidth]{low_tau_logtau4_theta90}
%	\caption{\textbf{Low optical thickness regimes}.  
%		Here we fix $\tauh=10^4$. 
%		The six panels correspond to $\theta=90^\circ$.
%		We vary \vrot increasing from left to right and \vout increasing
%		from top to bottom. 
%		The thin black line corresponds to the \lya line obtained with
%		CLARA without any rotation and the indicated outflow velocity.
%		The thick black line corresponds to the results including both
%		outflows and rotation.
%		The thick gray line shows the results of modifying the pure outflow
%		solution by the Doppler shift presented in Equation \ref{eq:shift_x}
%		(in thin line), if there is a radiative transfer of rotation and outflows
%		(thick and clear line), and if there is a radiative transfer of
%		only outflows, but also a Doppler shift from the rotational
%		velocity (thick and dark line).  
%		\label{fig:low_tau}}
%\end{figure}

@@

Addressing the second question we posed in the introduction (\emph{to what
extent is it possible to constrain rotational kinematics from the \lya
emission line?})  we can say that  the most straightforward approach
takes into account how the \lya spectra from two different sides of a
galaxy can detect approaching/receding gas motions
\citep{Prescott14,2018MNRAS.473.3907A}. 
In that case, we also show that one has to be careful and take into
account that the  peaks of these two spectra is smaller than a naively
expected value of $2\times\vrot$ due to the different weights at the
surface of the emitting regions.

As an application to observational data we use data from MUSE
\citep{2017A&A...606A..12H} to show to what extent the outputs of the
RT results presented here can be expected to change the results of
pure outflow simulations and thus bettwe match hundreds of
observed \lya spectra.
We find that the effect on general line morphology statistics
(standard deviation and skewness) is small, although a full constraint
of rotational properties certainly requires a full Markov Chain Monte
Carlo exploration of parameter space.

Our results suggest that rotation should also be considered as an
explanation of the non-vanishing \lya intensity at the line's center,
and become a test for rotation in the observed systems, provided that
the spectral resolution of the instrument is smaller than the expected
rotational velocities.

To summarize, our work confirms that the semi-analytic model based on
the Doppler Shift offers an easy-to-implement approximation to explore
such influence into already existing RT simulations and provide
stronger footing to the interpretation of present and future \lya
kinematics maps \citep[e.g][]{2018MNRAS.473.3907A}.

@@

\newpage
\appendix

@@

\begin{figure}
  \begin{center}
    \includegraphics[width=0.49\textwidth]{doppler_shift_logtau5_theta90}
  \end{center}
  \caption{\textbf{Qualitative trends of changing outflow and
      rotational velocity.}
    Same layout as Figure \ref{fig:doppler_shift},
    this time  $\tauh=10^5$ and $\theta=90^\circ$.}
\end{figure}

\begin{figure}
  \begin{center}
    \includegraphics[width=0.49\textwidth]{doppler_shift_logtau7_theta90}
  \end{center}
  \caption{\textbf{Qualitative trends of changing outflow and
      rotational velocity.}
    Same layout as Figure \ref{fig:doppler_shift},
    this time  $\tauh=10^7$ and $\theta=90^\circ$.}
\end{figure}

>>>>

\begin{abstract}
Outflows and rotation are two ubiquitous kinematic features in the gas
kinematics of galaxies. 
Here we introduce a semi-analytic model to quantify how rotating
outflows impact the morphology of the Lyman-$\alpha$ emission line.   
The model is contrasted against Monte Carlo radiative transfer
simulations of outflowing gas with additional solid body rotation.
We explore a range of neutral Hydrogen optical depth of
$10^5\leq\tauh\leq10^7$, rotational velocity $0\leq \vrot/\kms \leq
100$ and outflow velocity $0\leq \vout/\kms\leq 50$.  
We find three consequences of rotation.
First, it introduces a dependency with viewing angle; second it
produces a line broadening and third it increases the flux at the
line's center.
For all viewing angles, the semi-analytic model reproduces the
radiative transfer results for the line width and flux change at the
line's center within a $7\%$ and $50\%$ precision for an optical depth of
$\tauh=10^5$, respectively, and within $2\%$ and $1\%$ for an optical
depth of $\tauh=10^7$.
Using this toy model we also show that the peaks of integrated spectra
taken from opposite sides of an edge-on rotating gas distribution
should have a separation of $\frac{1}{2}\vrot$. 
The semi-analytic model presented here is a convenient tool to
introduce rotational kinematics as a post-processing step of idealized
Monte Carlo simulations, providing a framework to interpret \lya
spectra in systems where rotation is expected, or directly measured
through kinematic maps.  
\end{abstract}

@@

However, the gas dynamics in Lyman Alpha Emitter (LAE) galaxies are
more complex than pure rotation.
In many observations the \lya line profile has a single peak
redwards from the line's center, in other cases there is a double peak
but the peak on the red side is stronger
\citep[e.g.][]{2010ApJ...717..289S,Erb14,Trainor16}.   
These features have been explained as the consequence of multiple
\lya photon scatterings through a homogeneous outflowing shell of
neutral Hydrogen
\citep{Verhamme06,Orsi12,2015ApJ...812..123G}.

@@

@@

\subsection{\lya Kinematic Maps}
\label{subsec:kinematic}

Current observational facilities have the capability of spatially
resolving the extent of a LAE.
For instance \cite{Prescott14} presented observational results of a
Doppler shift when taking spectra at two opposite sides of a large
($\approx 80$kpc) LAE.
In more recent work \cite{2018MNRAS.473.3907A} mapped \lya emission
around a quasar.
In this case one could measure the displacement between the
peaks in the two different spectra. 
The interest of this test is that the location of the peaks only
depends on the kinematic properties of the sources and does not
change by convolution with a gaussian window function due to limited
instrumental resolution, as discussed in the previous section.

@@

In this toy model the distance between the peaks of the
receding/approaching spectra is close to $\sim 25\ \kms$, which is
a fourth of the naively expected value of $2\vrot=100\ \kms$,
due to the fact that only a small fraction of the photons are emitted
at the extreme of the galaxy having the maximum rotational velocity of
$50\ \kms$. 
Although it is a good approximation to think the rotating spectra by a
sum of Doppler shifts, the peak of the spectra is also weighted by the
amount of mass with a given line-of-sight velocity.

Spectrographs like the Multi Unit Spectroscopic Explorer (MUSE) could
obtain kinematic information from large samples of LAEs to build
velocity maps in Ly$\alpha$.  
This could be a natural extension of the work reported by
\cite{Herenz2016} on the velocity maps of several LARS (Lyman Alpha
Reference Sample) galaxies.  
The interpretation of such data could take into account the insights
and trends we have presented in this paper.

@@

Addressing the second question we posed in the introduction (\emph{to what
extent is it possible to constrain rotational kinematics from the \lya
emission line?})  we can say that  the most straightforward approach
takes into account how the \lya spectra from two different sides of a
galaxy can detect approaching/receding gas motions
\citep{Prescott14,2018MNRAS.473.3907A}. 
In that case, we also show that one has to be careful and take into
account that the  peaks of these two spectra is smaller than a naively
expected value of $2\times\vrot$ due to the different weights at the
surface of the emitting regions.   
Here the semi-analytic model also provides the correct quantitative
insight.

To summarize, our work confirms that the semi-analytic model based on
the Doppler Shift offers an easy-to-implement approximation to explore
such influence into already existing radiative transfer simulation
simulations and provide stronger footing to the interpretation of
present and future \lya kinematics maps
\citep[e.g][]{2018MNRAS.473.3907A}.

@@

\appendix

@@

\begin{figure*}
  \begin{center}
    \includegraphics[width=0.49\textwidth]{doppler_shift_logtau5_theta90}
    \includegraphics[width=0.49\textwidth]{doppler_shift_logtau7_theta90}
  \end{center}
  \caption{\textbf{Qualitative trends of changing outflow and
      rotational velocity.}
    Same layout as Figure \ref{fig:doppler_shift}. On the left:
    $\tauh=10^5$ and $\theta=90^\circ$; on the right:     this time  $\tauh=10^7$ and $\theta=90^\circ$.}
\end{figure*}
